Viscous Transport in Eroding Porous Media

### 2019-08-21 docs/transport/erosion_transport.tex
working on intro and other sections

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow in many natural occurring geometries involves flows that vary over
many orders of magnitude.  For example, the length scales of a porous
media can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}.  Porous media flow plays an
important role in several environmental and industrial applications.
Therefore, understanding the dynamics of pore-scale flow is central to
upscale the governing fluid equations, characterize
dispersion~\cite{saf1959}, and quantify mixing. However, because of the
large disparity of length and velocity scales, constitutive laws that
link the microscopic scale to macroscopic scale must be
developed~\cite{mil-chr-imh-mcb-ped1998}.  Therefore, to find closure
formulas, accurate simulations at microscopic scales must be performed
and analyzed.  A few examples of closure formulas and coarse-grained
variables include
pressure-saturation-permeability~\cite{mil-chr-imh-mcb-ped1998}, first
passage time distribution~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosity~\cite{hak-com-den2019,
matyka2008tortuosity, dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, and anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}.

@@

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  The incompressible
Stokes equation are solved by first recasting them as a boundary
integral equation (BIE).  The BIE is discretized with a high-order
quadrature method that resolves nearly touching bodies with a
near-singular integration scheme~\cite{bar-wu-vee2015}.  The erosion
dynamics are simulated by applying a stable second-order time stepping
Runge-Kutta method to the {\thL} formulation~\cite{hou-low-she1994} of
the eroding bodies.  Given the solution of the incompressible Stokes
equations are solved in an eroded porous geometry, we investigate
different metrics to characterize transport.  We focus on three metrics:
the tortuosity, the anomalous dispersion, and the distribution of
geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
dispersion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
\end{align}
where $\uu$ is the fluid velocity.  Since there is no diffusion, all
dispersion is a result of the particle spreading induced by the porous
geometry.

@@

The tortuosity is a first statistical moment of the tracer trajectories.
However, to measure spread of trajectories, second-order moments are
computed. By computing the variance of the lengths of trajectories in a
porous media, non-local anomalous dispersion is
observed~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}, resulting in super-diffusive dispersion.
Anomalous dispersion has been investigated for a variety of porous
geometries, but the effect of erosion on the anomalous dispersion rate
is lacking.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
\begin{itemize}
  \item Use novel near-singular Barycentric integration scheme for
    Stokes equation outlined in~\cite{bar-wu-vee2015}. (Remember to
    include historically significant reference Shelley
    \cite{baker1986boundary}).  Also need discuss Helsing and Ojala who
    did a similar panel-based quadrature~\cite{hel-oja2008}
  \item Need to compute one additional derivative since we need the
    derivative of the velocity to find the vorticity and stress tensor
  \item Annealing algorithm to create initial body configurations of specified area fraction specified size distribution (NM can do the writing for this).
  \item Tracer trajectories at different time steps
  \item Compute the anomalous dispersion rate using similar techniques
  as the paper with Pietro~\cite{dea-qua-bir-jua2018}
  \item Need a novel way to do time stepping through the complex
    geometry
\end{itemize}

@@

\begin{itemize}
\item Porous aquifers are a good application of our work

\item \cite{dar-mcc1998} show that if there is a region with high
porosity, this greatly affects the relationships between the porosity
and permeability.  Uses a Lattice-Boltzmann method.

\item Continuous Time Random Walks~\cite{ber-cor-den-sch2006}

@@

\item Near-singular integration such as regularization, QBX, Lagrange
interpolation, panel-based quadrature, etc.

\item The classic work of Carman~\cite{car1937} reviews early work on
granular flow.

@@

\item \cite{won-kop-tom1984} investigates relationships between the
porosity of a rock to its permeability, surface area, and the electrical
conductivity of salt water within the rock.  They show that the pore
size distribution influences both the conductivity and permeability.

\item In~\cite{rot1988}, Rothman applies a cellular automaton method to
simulate flow in porous media.  The method is probabilistic and agrees
well with analytic solutions.  However, the resulting flow is not
smooth, so quantities such as the shear stress is challenging to stably
compute.

\item In~\cite{leb-den-car2008}, a statistical model is used to describe
transport in a high heterogeneous porous media.

\item In~\cite{bar2014}, a local expansion technique similar to QBX is
used.  The advantage is that this extends naturally to 3D.  Also,
Barnett analyzed the error of the Laplace DLP and SLP and compared the
numerical results to the theory.

\item Suite of near-singular integration methods such as QBX,
panel-based quadrature, other local interpolants, regularization.

\item \cite{ioa-pap-per1991} were the first to develop quadrature for
analytic functions in the interior case.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this section, we apply our new quadrature method to simulate erosion
of several different configurations of eroding bodies.  The fluid domain
is $\Omega$ whose outer solid walls is a rectangle $[-3,3] \times
[-1,1]$ with rounded off corners.  We only place bodies only in $[-1,1]
\times [-1,1]$.  For all simulations, we discretize $\Gamma = \bd
\Omega$ with $N_\out=1024$ points.  For all but the first example, the
Dirichlet boundary condition on $\Gamma$ is a hagen-poiseuille flow.  As
the bodies erode, the flow rate is adjusted to maintain the physical
constant pressure drop boundary condition.  To maintain this pressure
drop, we compute the pressure at near the inlet and outlet at each time
step as described in our previous work~\cite{qua-moo2018}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow in many natural occurring geometries involves flows that vary over
many orders of magnitude.  For example, the length scales of a porous
media can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}.  Porous media flow plays an
important role in several environmental and industrial applications.
Therefore, understanding the dynamics of pore-scale flow is central to
upscale the governing fluid equations, characterize
dispersion~\cite{saf1959}, and quantify mixing. However, because of the
large disparity of length and velocity scales, constitutive laws that
link the microscopic scale to macroscopic scale must be
developed~\cite{mil-chr-imh-mcb-ped1998}.  Therefore, to find closure
formulas, accurate simulations at microscopic scales must be performed
and analyzed.  A few examples of closure formulas and coarse-grained
variables include
pressure-saturation-permeability~\cite{mil-chr-imh-mcb-ped1998},
permeability-porosity~\cite{dar-mcc1998, car1937, won-kop-tom1984},
first passage time distribution~\cite{ber-sch-sil2000,
hym-den-hag-kan2019, cve-che-wen1996}, tortuosity~\cite{hak-com-den2019,
matyka2008tortuosity, dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, and anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}.

@@

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas model~\cite{kop-kat-tim1996}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
dispersion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no diffusion, all
dispersion is a result of the particle spreading induced by the porous
geometry.  Since there is no diffusion, we apply a high-order explicit
Runge-Kutta time stepping method to compute tracer trajectories.
However, other groups have simulated trajectories using continuous time
random walks~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006} and a Fokker-Planck equation~\cite{ica-den2019}.

@@

The tortuosity is a first statistical moment of the tracer trajectories.
However, to measure spread of trajectories, second-order moments are
computed. By computing the variance of the lengths of trajectories in a
porous media, non-local anomalous dispersion is
observed~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}, resulting in super-diffusive dispersion.
The dispersion depends on the amount of time trajectories spend in low
and high velocity regimes~\cite{ber-sch2001}.  Anomalous dispersion has
been investigated for a variety of porous geometries, but not in the
context of eroding porous geometries.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
\begin{itemize}
  \item Use novel near-singular Barycentric integration scheme for
    Stokes equation outlined in~\cite{bar-wu-vee2015}. (Remember to
    include historically significant reference Shelley
    \cite{bak-she1986}).
  \item Need to compute one additional derivative since we need the
    derivative of the velocity to find the vorticity and stress tensor
  \item Annealing algorithm to create initial body configurations of
  \item Tracer trajectories at different time steps
  \item Compute the anomalous dispersion rate using similar techniques
  as the paper with Pietro~\cite{dea-qua-bir-jua2018}
  \item Need a novel way to do time stepping through the complex
    geometry
\end{itemize}

@@

\begin{itemize}
\item \cite{dar-mcc1998} show that if there is a region with high
porosity, this greatly affects the relationships between the porosity
and permeability.  Uses a Lattice-Boltzmann method.

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this section, we apply our new quadrature method to simulate erosion
of several different configurations of eroding bodies.  The fluid domain
is $\Omega$ whose outer solid walls is a rectangle $[-3,3] \times
[-1,1]$ with rounded off corners.  We initialize the bodies randomly in
$[-1,1] \times [-1,1]$ using a randomized method that chooses a random
center and radii, and then accepts the new body if it is sufficiently
separated from all other bodies.  Since the error of our quadrature
method is bounded independent of the geometry, we consider bodies that
are separated by less than 10\% of an arclength spacing.  For all
simulations, we discretize $\Gamma = \bd \Omega$ with $N_\out=1024$
points.  For all but the first example, the Dirichlet boundary condition
on $\Gamma$ is a hagen-poiseuille flow.  As the bodies erode, the flow
rate is adjusted to maintain the physical constant pressure drop
boundary condition.  To maintain this pressure drop, we compute the
pressure at near the inlet and outlet at each time step as described in
our previous work~\cite{qua-moo2018}.
